---
title: "Eddie module stream discharge from handout"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


Libraires
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# load the USGS package
library(dataRetrieval)
```

Data from Mississippi River at St Louis Site 07010000
https://waterdata.usgs.gov/mo/nwis/uv/?site_no=07010000&PARAmeter_cd=00065,63160,00060x

Read in file for ...
```{r read in file}
# Enter the site number below
# Leave the start data and end data open as it will retreive all data
siteNo <-         "07010000"
start.date <-     "1800-01-01"
end.date <-       "2050-01-01"

peak_q.df <- readNWISpeak(
              siteNumbers = siteNo,
              startDate = start.date,
              endDate = end.date,
              asDateTime = TRUE,
              convertType = TRUE)
  
# Clean up the names and remove extra data
peak_q.df <- peak_q.df %>% 
  select(agency_cd, 
         site_no, 
         date = peak_dt, 
         discharge_cfs = peak_va, 
         gage_ht_ft = gage_ht)
```

# Plot yearly data in timeline
```{r}
q.plot <- neversink_peak.df %>% 
  ggplot(aes(date, discharge_cfs)) +
  geom_point() +
  geom_line()
q.plot
```

# Look at the highest discharge on record
```{r}
peak_q.df %>% 
  filter(
    discharge_cfs  == max(discharge_cfs, na.rm=TRUE))
```

Discharge Statistics
```{r}
peak_q.df %>%
  summarize(
    number = sum(!is.na(discharge_cfs)),
    mean_discharge_csf = mean(discharge_cfs, na.rm=TRUE),
    std_dev_discharge_csf = sd(discharge_cfs, na.rm=TRUE),
    stderr = sd(discharge_cfs, na.rm=TRUE)/sqrt(sum(!is.na(discharge_cfs)))
  )
```

# Calculating probability
```{r}
floodprob.df <- peak_q.df %>%
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

# Recurrance interval plot
```{r}
floodprob.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs)) +
  geom_point() +
  scale_y_continuous(labels= comma )
```

# Log Recurrance interval plot
```{r}
floodprob.df %>% 
  ggplot(aes(x = log10(recurrence), y=log10(discharge_cfs))) +
  geom_point()
```





