---
title: "Eddie Module on Stream discharge"
output: html_notebook
---



# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


Libraires
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# load the USGS package
library(dataRetrieval)
```


USGS 12120000 MERCER CREEK NEAR BELLEVUE, WA
https://waterdata.usgs.gov/nwis/inventory/?site_no=12120000

USGS 12113000, Green River near Auburn, Washington
https://waterdata.usgs.gov/nwis/inventory/?site_no=12113000

USGS 12108500, Newaukum Creek near Black Diamond, Washington
https://waterdata.usgs.gov/nwis/inventory/?site_no=12108500


# Read in file from excel sheet
```{r}
green.df <- read_excel("../data/Green R.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Green")
mercer.df <- read_excel("../data/Mercer Cr.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Mercer")
new.df <- read_excel("../data/Newaukum Cr.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Newaukum")
```

# Combine data
```{r}
streams.df <- bind_rows(green.df, mercer.df, new.df)
```


# Graph of yearly peak discharge
```{r}
streams.df %>% 
  ggplot(aes(x = date, y=discharge_cfs, color=site)) +
  geom_line() +
  geom_point()
```

# Look at the highest discharge on record
```{r}
streams.df %>%
  group_by(site) %>% 
  filter(
    discharge_cfs  == max(discharge_cfs, na.rm=TRUE))
```

Discharge Statistics
```{r}
streams.df %>%
  group_by(site) %>% 
  summarize(
    number = sum(!is.na(discharge_cfs)),
    mean_discharge_csf = mean(discharge_cfs, na.rm=TRUE),
    std_dev_discharge_csf = sd(discharge_cfs, na.rm=TRUE),
    stderr = sd(discharge_cfs, na.rm=TRUE)/sqrt(sum(!is.na(discharge_cfs)))
  )
```

# Calculating probability
```{r}
floodprob.df <- streams.df %>%
  group_by(site) %>% 
  arrange(site, desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
  
```

# Recurrance interval plot
```{r}
floodprob.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=site)) +
  geom_point() +
  scale_y_continuous(labels= comma )
```

# Log Recurrance interval plot
```{r}
floodprob.df %>% 
  group_by(site) %>% 
  ggplot(aes(x = log10(recurrence), y=log10(discharge_cfs), color=site)) +
  geom_point()
```


# Look at Mercer Creek Before and after 1977
```{r} 
mercer_urban.df <- mercer.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1977 ~ "pre_1977",
    year > 1977 ~ "post_1977"
  ))
```

# Calculating probability Mercer Cr Pre Post 1977
```{r}
mercer_urban.df <- mercer_urban.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

# Mercer Creek pre post recurrece interval plot
```{r}
mercer_urban.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )

```


# Green River Dam Effects
```{r} 
green_dam.df <- green.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1961 ~ "pre_1961",
    year > 1961 ~ "post_1961"
  ))
```

# Calculating probability Green R Pre Post 1961
```{r}
green_dam.df <- green_dam.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

# Green R pre post recurrece interval plot
```{r}
green_dam.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )

```

# New R River Dam Effects
```{r} 
new_dummy.df <- new.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1977 ~ "pre_1977",
    year > 1977 ~ "post_1977"
  ))
```

# Calculating probability Green R Pre Post 1961
```{r}
new_dummy.df <- new_dummy.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```



# Mercer Creek pre post recurrece interval plot
```{r}
new_dummy.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )
```





