---
title: "Eddie module stream discharge from handout"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


Libraires
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# load the USGS package
library(dataRetrieval)
```

current data
https://waterdata.usgs.gov/ny/nwis/uv/?site_no=01435000&PARAmeter_cd=00065,00060,63160

all data
https://waterdata.usgs.gov/nwis/inventory/?site_no=01435000&agency_cd=USGS


Read in file for ...
```{r read in file}

# The site is Neversink and the site number is 01435000
# USGS 01435000 NEVERSINK RIVER NEAR CLARYVILLE NY

site_no <-          "01435000"
par_code <-         "00060"
start.date <-      "1800-01-01"
end.date <-        "2050-01-01"

neversink_daily.df <- readNWISdv(siteNumbers = site_no,
                        parameterCd = par_code,
                        startDate = start.date,
                        endDate = end.date)
  
# rename the columns
neversink_daily.df <- renameNWISColumns(neversink_daily.df)
```

# we need to make a date column to plot with
```{r}
neversink_daily.df <- neversink_daily.df %>%
  select(agency_cd, 
         site_no, 
         date = Date, 
         discharge_cfs = Flow) 
```

# we need to make a date column to plot with
```{r}
neversink_daily.df <- neversink_daily.df %>%
  mutate(month = month(date))
```



# Plot all data
```{r}
q.plot <- neversink_daily.df %>% 
  ggplot(aes(date, discharge_cfs)) +
  geom_point() +
  geom_line()
q.plot
```

# Plot February data 
```{r}
feb.plot <- neversink_daily.df %>% 
  filter(month == 2) %>%
  ggplot(aes(date, discharge_cfs)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
feb.plot
  
```

We can make it interactive
```{r}
ggplotly(feb.plot)
```


August data
```{r}
aug.plot <- neversink_daily.df %>%
  filter(month == 8) %>%
  ggplot(aes(date, discharge_cfs)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
aug.plot
```

# Compare the plots
```{r}
feb_aug.plot <- feb.plot +
                aug.plot +
                plot_layout(ncol =1)
feb_aug.plot
```

# add trend lines as straight lines
```{r}
feb_aug.plot <- feb.plot + geom_smooth(method="lm") +
                aug.plot + geom_smooth(method="lm") +
                plot_layout(ncol =1)
feb_aug.plot
```

# How to see regression statistics
```{r}
feb.model <- neversink_daily.df %>% 
  filter(month == 2) %>%
  lm(discharge_cfs ~ date, data=.)
summary(feb.model)

```


```{r}
aug.model <- neversink_daily.df %>% 
  filter(month == 8) %>%
  lm(discharge_cfs ~ date, data=.)
summary(aug.model)

```


