---
title: "Eddie module stream discharge from handout"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


Libraires
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# load the USGS package
library(dataRetrieval)
```

Data from:
USGS 12120000 MERCER CREEK NEAR BELLEVUE, WA
https://waterdata.usgs.gov/nwis/inventory/?site_no=12120000

USGS 12113000, Green River near Auburn, Washington
https://waterdata.usgs.gov/nwis/inventory/?site_no=12113000

USGS 12108500, Newaukum Creek near Black Diamond, Washington
https://waterdata.usgs.gov/nwis/inventory/?site_no=12108500


Read in file for ...
```{r read in file}
# Enter the site number below
# Leave the start data and end data open as it will retreive all data
siteNo <-         c("12120000", "12113000", "12108500")
start.date <-     "1800-01-01"
end.date <-       "2050-01-01"

peak_q.df <- readNWISpeak(
              siteNumbers = siteNo,
              startDate = start.date,
              endDate = end.date,
              asDateTime = TRUE,
              convertType = TRUE)
  
# Clean up the names and remove extra data
peak_q.df <- peak_q.df %>% 
  select(agency_cd, 
         site_no, 
         date = peak_dt, 
         discharge_cfs = peak_va, 
         gage_ht_ft = gage_ht)
```

Rename the sites
```{r}
peak_q.df <- peak_q.df %>% 
  mutate(site = case_when(
    site_no == 12120000 ~ "Mercer Cr",
    site_no == 12113000 ~ "Green R",
    site_no == 12108500 ~ "Newaukum Cr",
    TRUE ~ "other"
  ))
```

# Subdivide the sites
```{r}
green.df <- peak_q.df %>% filter(site == "Green R")
mercer.df <- peak_q.df %>% filter(site == "Mercer Cr")
new.df <- peak_q.df %>% filter(site == "Newaukum Cr")
```


# Graph of yearly peak discharge
```{r}
peak_q.df %>% 
  ggplot(aes(x = date, y=discharge_cfs, color=site)) +
  geom_line() +
  geom_point()
```

# Look at the highest discharge on record
```{r}
peak_q.df %>%
  group_by(site) %>% 
  filter(
    discharge_cfs  == max(discharge_cfs, na.rm=TRUE))
```

Discharge Statistics
```{r}
peak_q.df %>%
  group_by(site) %>% 
  summarize(
    number = sum(!is.na(discharge_cfs)),
    mean_discharge_csf = mean(discharge_cfs, na.rm=TRUE),
    std_dev_discharge_csf = sd(discharge_cfs, na.rm=TRUE),
    stderr = sd(discharge_cfs, na.rm=TRUE)/sqrt(sum(!is.na(discharge_cfs)))
  )
```

# Calculating probability
```{r}
floodprob.df <- peak_q.df %>%
  group_by(site) %>% 
  arrange(site, desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
  
```

# Recurrance interval plot
```{r}
floodprob.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=site)) +
  geom_point() +
  scale_y_continuous(labels= comma )
```

# Log Recurrance interval plot
```{r}
floodprob.df %>% 
  group_by(site) %>% 
  ggplot(aes(x = log10(recurrence), y=log10(discharge_cfs), color=site)) +
  geom_point()
```


# Look at Mercer Creek Before and after 1977
```{r} 
mercer_urban.df <- mercer.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1977 ~ "pre_1977",
    year > 1977 ~ "post_1977"
  ))
```

# Calculating probability Mercer Cr Pre Post 1977
```{r}
mercer_urban.df <- mercer_urban.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

# Mercer Creek pre post recurrece interval plot
```{r}
mercer_urban.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )

```


# Green River Dam Effects
```{r} 
green_dam.df <- green.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1961 ~ "pre_1961",
    year > 1961 ~ "post_1961"
  ))
```

# Calculating probability Green R Pre Post 1961
```{r}
green_dam.df <- green_dam.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

# Green R pre post recurrece interval plot
```{r}
green_dam.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )

```

# New R River Dam Effects
```{r} 
new_dummy.df <- new.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1977 ~ "pre_1977",
    year > 1977 ~ "post_1977"
  ))
```

# Calculating probability Green R Pre Post 1961
```{r}
new_dummy.df <- new_dummy.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```



# Mercer Creek pre post recurrece interval plot
```{r}
new_dummy.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )
```



