---
title: "Eddie module stream discharge from handout"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


Libraires
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# load the USGS package
library(dataRetrieval)
```

current data
https://waterdata.usgs.gov/ny/nwis/uv/?site_no=01435000&PARAmeter_cd=00065,00060,63160

all data
https://waterdata.usgs.gov/nwis/inventory/?site_no=01435000&agency_cd=USGS


Read in file for ...
```{r read in file}

# code is from 
# 

# The site is Neversink and the site number is 01435000
# USGS 01435000 NEVERSINK RIVER NEAR CLARYVILLE NY

siteNo <-     "01435000"
pCode_monthly <-      "00060"
start.date <- "1937-11-01"
end.date <-   "2019-12-25"

neversink_monthly.df <-   readNWISstat(siteNumbers = siteNo,
                        parameterCd = pCode_monthly,
                        startDate = start.date,
                        endDate = end.date,
                        statReportType="daily",
                        statType="mean")
  
# rename the columns
neversink_monthly.df <- renameNWISColumns(neversink_monthly.df)

names(neversink_monthly.df)
```

# we need to make a date column to plot with
```{r}
neversink_monthly.df <- neversink_monthly.df %>%
  mutate(
    date = ymd(paste(year_nu, month_nu, "01", sep="-"))
  )

```



# Plot all data
```{r}
all.plot <- ggplot(neversink_monthly.df, aes(date, mean_va)) +
  geom_point() +
  geom_line()
all.plot
```



# Plot February data 

```{r}
feb.plot <- neversink_monthly.df %>%
  filter(month_nu == 2) %>%
  ggplot(aes(date, mean_va)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
feb.plot
  
```

We can make it interactive
```{r}
ggplotly(feb.plot)
```


August data
```{r}
aug.plot <- neversink_monthly.df %>%
  filter(month_nu == 8) %>%
  ggplot(aes(date, mean_va)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
aug.plot
```

# Compare the plots
```{r}
feb_aug.plot <- feb.plot +
                aug.plot +
                plot_layout(ncol =1)
feb_aug.plot
```

# add trend lines as straight lines
```{r}
feb_aug.plot <- feb.plot + geom_smooth(method="lm") +
                aug.plot + geom_smooth(method="lm") +
                plot_layout(ncol =1)
feb_aug.plot
```

# How to see regression statistics
```{r}
feb.model <- neversink_monthly.df %>% 
  filter(month_nu == 2) %>%
  lm(mean_va ~ date, data=.)
summary(feb.model)

```


```{r}
aug.model <- neversink_monthly.df %>% 
  filter(month_nu == 8) %>%
  lm(mean_va ~ date, data=.)
summary(aug.model)

```



# Try another site
Use this site to find a new location
https://maps.waterdata.usgs.gov/mapper/index.html

```{r}
# The site is Neversink and the site number is 01435000
# USGS SAGAVANIRKTOK R NR PUMP STA 3 AK

siteNo_monthly <-     "15908000"
pCode_monthly <-      "00060"
start.date_monthly <- "1982-01"
end.date_monthly <-   "2019-08"

sag_monthly.df <-   readNWISstat(siteNumbers = siteNo_monthly,
                        parameterCd = pCode_monthly,
                        startDate = start.date_monthly,
                        endDate = end.date_monthly,
                        statReportType="Monthly",
                        statType="mean")
  
# rename the columns
sag_monthly.df <- renameNWISColumns(sag_monthly.df)

names(sag_monthly.df)
```

# we need to make a date column to plot with
```{r}
sag_monthly.df <- sag_monthly.df %>%
  mutate(
    date = ymd(paste(year_nu, month_nu, "01", sep="-"))
  )

```


# Plot February data 

```{r}
feb_new.plot <- sag_monthly.df %>%
  filter(month_nu == 2) %>%
  ggplot(aes(date, mean_va)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
feb_new.plot
  
```



August data
```{r}
aug_new.plot <- sag_monthly.df %>%
  filter(month_nu == 8) %>%
  ggplot(aes(date, mean_va)) +
  geom_point() +
  geom_line() +
  labs(y="Mean Monthly Discharge m^3/sec", x="Date")
aug_new.plot
```


# Compare the plots
```{r}
feb_aug_new.plot <- feb_new.plot +
                aug_new.plot +
                plot_layout(ncol =1)
feb_aug_new.plot
```




# Never sink Peak Flows
```{r}
siteNo_peak <-     "01435000"
pCode_peak <-      "00060"
start.date_peak <- "1850-01-01"
end.date_peak <-   "2019-08-02"

neversink_peak.df <-   readNWISpeak(siteNumbers = siteNo_peak)

# rename the columns
neversink_peak.df <- renameNWISColumns(neversink_peak.df)

names(neversink_peak.df)

```



# Plot the data

```{r}
peak_flow.plot <- neversink_peak.df  %>%
  ggplot(aes(peak_dt , peak_va)) + 
  geom_point() +
  geom_line()
ggplotly(peak_flow.plot)
```

# mean and std flows
```{r}
skim(neversink_peak.df)
```

# calculate flood probability
this is one way to do it but is not the best
```{r}

neversink_floodfreq.df <- neversink_peak.df %>%
  select(agency_cd, site_no, peak_dt, peak_va, gage_ht) %>%
  arrange(desc(peak_va)) %>% 
  mutate(
    year =  year(peak_dt)
  ) %>%
  group_by(year) %>%
  filter(peak_va == max(peak_va)) %>%
  ungroup() %>%
  mutate(rank = dense_rank(desc(peak_va)))

```


```{r calculate probability}

neversink_floodfreq.df <- neversink_floodfreq.df %>%  
  ungroup() %>% 
  mutate(total_n = n()) %>% 
  rowwise() %>%
  mutate(probability = rank/(total_n+1))

```


```{r calculate long term probability}
neversink_floodfreq.df <- neversink_floodfreq.df %>%  
  mutate(
    recurrence_inteval = 1/probability
  )


```



```{r}
neversink_floodfreq.df %>%
  ggplot(aes(log(recurrence_inteval), log(peak_va))) + 
  geom_point()


```

