---
title: "Eddie Module on Stream discharge excel based data"
output:
  html_document:
    df_print: paged
---

# Excel based analysis of discharge and urbanization

This uses the excel based data for the analysis of the effects of urbanization 


## Install libraries 
The libraries need to be installed only one time and then loaded each time the script is opened and run.  
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("devtools") # essential in installing other thigs
# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("lubridate") # makes working with dates easier
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots
# install.packages(skimr)
# devtools::install_github("thomasp85/patchwork") # multiple plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method
```


## Load Libraires
Load the libraries each time you run the script.    
```{r laod libraries, message=TRUE, warning=TRUE}
# load these every time  
library(tidyverse)
library(scales)
library(lubridate)
library(janitor)
library(readxl)
library(skimr)
library(plotly)
library(patchwork)

# # load the USGS package
# library(dataRetrieval)
```


## USGS Sites       
These are the sites that are used in the data analysis below. The data can be downloaded again and then used for the analsyses.     

USGS 12120000 MERCER CREEK NEAR BELLEVUE, WA       
https://waterdata.usgs.gov/nwis/inventory/?site_no=12120000      
       
USGS 12113000, Green River near Auburn, Washington      
https://waterdata.usgs.gov/nwis/inventory/?site_no=12113000       
      
USGS 12108500, Newaukum Creek near Black Diamond, Washington     
https://waterdata.usgs.gov/nwis/inventory/?site_no=12108500     

## Read in Excel Files      
This reads in the files into 3 separate dataframes that appear in the Environment tab in the upper right.    
```{r read files}

# This reads in the Green River data
green.df <- read_excel("../data/Green R.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Green") # adds Green to site name

# Read in Mercer Cr data
mercer.df <- read_excel("../data/Mercer Cr.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Mercer") # adds Mercer to site name


# read in the Newaukum data
new.df <- read_excel("../data/Newaukum Cr.xlsx",
                 guess_max = 10000) %>% 
  mutate(site="Newaukum") # adds Newaukum to site name


```

## Combine all data frames      
This will merge all the data files in a linear manner.   The bind_rows command is nicer than the base R vesion as it will add all the data and if there are missing values in one dataframe it will still perform the merge.     

```{r bind_rows}
streams.df <- bind_rows(green.df, mercer.df, new.df)
```


## Graph of yearly peak discharge
This graphs the discharge data from each stream. This is the max flow that occured each year. If the name of a variable is put inside of the aes( here ) statment it will map the values of that vaiable to the type of symbol or color you have designated.   These commands may be:
`aes(x = x_variable, y = y_variable, *color* = variable name of sites)`    
`aes(x = x_variable, y = y_variable, *fill* = variable name of sites)`    
`aes(x = x_variable, y = y_variable, *shape* = variable name of sites)`    
`aes(x = x_variable, y = y_variable, *line_type* = variable name of sites)`        

This is the graph using the commands above and you can try changing these around <br>      

```{r graph of discharge}
streams.df %>% # use this data set
  ggplot(aes(x = date, y=discharge_cfs, color=site)) + # graph these variables with grouping based on site
  geom_line() + # add lines
  geom_point() # add points
```

## Table of Highest dischage    
There are several ways to obtian the maximum values.     
1) Filter the data that has been grouped based on sites. This will return the rows of the data based on that filter.      
2) Summarize the data. You coul do a grouping command of site and return the max discharge but this would require some extra code and would be a bit clunkier.      
3) Use the Skim package to get a summary of the data by groups     

```{r max discharge}
streams.df %>% # use this data set
  group_by(site) %>%  # group_by site
  filter(
    discharge_cfs  == max(discharge_cfs, na.rm=TRUE)) 
# filter the discharge_cfs column by the maximum discharge removing na values
```

## Discharge Statistics     
This will provide some basic summary statistics of the data.    

```{r discharge stats}
streams.df %>%  # use this data
  group_by(site) %>%  # group by site
  summarize( # provides summarization of data    
    number = sum(!is.na(discharge_cfs)), # provides a sum of the number of values that are not missing
    mean_discharge_csf = mean(discharge_cfs, na.rm=TRUE), # mean of discharge
    std_dev_discharge_csf = sd(discharge_cfs, na.rm=TRUE), # standard deviation
    stderr = sd(discharge_cfs, na.rm=TRUE)/sqrt(sum(!is.na(discharge_cfs))) # standard error and works with missing values
  )
```

# Calculating flood probability
This will calculate flood probability
```{r flood_probability}
floodprob.df <- streams.df %>% # store tehe results in the floodprob.df from the streams data
  group_by(site) %>%  # group by site
  arrange(site, desc(discharge_cfs))%>%  # sorts data by site and discharge largest to smallest (desc)
  mutate(rank  = 1:n()) %>%  # adds in a rank of discharge by site from largest to smallest
  mutate(prob = rank / (n()+1)) %>% # calculates probabliilty of the flood occurring
  mutate(pct_prob = prob * 100) %>%  # calculates percent probability
  mutate(recurrence  = 1/prob) # calculates recurrence interval
```

# Plot of Recurrence Interval        
Plots the recurrence interval for each of the sites.

```{r recurrence plot}
floodprob.df %>%  # use this data
  ggplot(aes(x = recurrence, y=discharge_cfs, color=site)) + # plot these variables grouped by sites
  geom_point() + # add points
  scale_y_continuous(labels= comma ) # make the y axis a continuous scale with commas to make easier to read
```

# Log Recurrence interval plot    

```{r log_recurrence plot}
floodprob.df %>%  # use this data
  group_by(site) %>%  # group by site again not necessary so why did I put it here - not sure
  ggplot(aes(x = log10(recurrence), y=log10(discharge_cfs), color=site)) + # plot these variables and group by site
  geom_point() # add points
```

# Look at data before and after particular times     
This will work withthe individual streams to look at changes before and after a particular time.    

## Mercer Creek Before and after 1977     
This will create a new column before and after 1977. This could be done as usual a number of different ways but this might be the easiest to work with.      
```{r mercer add column} 
mercer_urban.df <- mercer.df %>%  # create a new data frame from mercer.df
  mutate(year = year(date)) %>%  # create a year column with the year derived from date
  mutate(pre_post = case_when( # this is essentially and if_else statment.... so 
    year <= 1977 ~ "pre_1977",  # if year is less than or equal to 1977 add "pre_1977" to pre_post
    year > 1977 ~ "post_1977"   # if year is greater than 1977 add "post_1977" to pre_post
  ))
```

## Calculating probability Mercer Cr Pre Post 1977
Rather than grouping based on stream we are grouping based on pre_post 1977. This is essentially tyhe same thing we did above to caclulate probability.     
```{r}
mercer_urban.df <- mercer_urban.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

## Mercer Creek pre post recurrece interval plot 
Thsi plots the data as above grouped fro Mercer Cr pre and post 1977
```{r}
mercer_urban.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )
```


## Green River Dam Effects
This looks at the effect of a dam on Green River. Again this uses the same code above and since the variable names are the same we can copy the code form above to hear and change some of the values.  
```{r} 
green_dam.df <- green.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1961 ~ "pre_1961",
    year > 1961 ~ "post_1961"
  ))
```

## Calculating flood probability in Green R Pre Post 1961

```{r}
green_dam.df <- green_dam.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```

## Recurrece interval plot Green R pre post dam 

```{r}
green_dam.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )

```

## Newaukum River Dam Effects
```{r} 
new_dummy.df <- new.df %>% 
  mutate(year = year(date)) %>% 
  mutate(pre_post = case_when(
    year <= 1961 ~ "pre_1961",
    year > 1961 ~ "post_1961"
  ))
```

## Calculating flood probability Newaukum R Pre Post 1961
```{r}
new_dummy.df <- new_dummy.df %>%
  group_by(pre_post) %>% 
  arrange(desc(discharge_cfs))%>% 
  mutate(rank  = 1:n()) %>% 
  mutate(prob = rank / (n()+1)) %>%
  mutate(pct_prob = prob * 100) %>% 
  mutate(recurrence  = 1/prob)
```



## Newaukum R pre post recurrece interval plot
```{r}
new_dummy.df %>% 
  ggplot(aes(x = recurrence, y=discharge_cfs, color=pre_post)) +
  geom_point() +
  scale_y_continuous(labels = comma )
```





