---
title: "Project Eddie Stream Discharge Module"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# USGS has libraries to retreive data from their system.    

# install libraries - do only one time when first runnign script    
```{r install packages, message=TRUE, warning=TRUE}
# https://github.com/USGS-R/dataRetrieval

# general R packages to install if you dont have them
# remove the "#" if you need to run the installations

# install.packages("tidyverse") # installs a lot of things and ggplot
# install.packages("scales") # allows great scale formatting on ggplot
# install.packages("janitor") # clean names of columns and other things
# install.packages("readxl") # read in excel files
# install.packages("plotly") # interactive plots

# specific to this module
# install.packages("dataRetrieval") # USGS Data Retreiveal Method

```

We need to load the libraries in these packages - this has to be done evertime    
```{r load libraries}
# load these every time  
library(tidyverse)
library(scales)
library(janitor)
library(readxl)
library(plotly)

# load the USGS package
library(dataRetrieval)

```

# Lets try to get some data 
The codes you will get are belowshort
Parameter Codes	  Short Name     Statistic Codes    Short name
     00060	      Discharge         00001            Maximum
     00065	      Gage Height       00002            Minimum
     00010	      Temperature       00003            Mean
     00400	      pH                00008            Median
```{r}
# Create variables to retreive
siteNo <-     "01435000"         # this is the site code
pCode <-      "00060"             # this is the parameter(s) to retreive
start.date <- "1992-10-01"  # Begining date to retreive
end.date <-   "2019-08-02"     # End date to retreive

# the variables above are entered below 
neversink.df <- readNWISuv(siteNumbers = siteNo,
                          parameterCd = pCode,
                          startDate = start.date,
                          endDate = end.date)
```


Explore the data
```{r}
# the names in USGS format
names(neversink.df)

```

# Rename variables
```{r}
neversink.df <- renameNWISColumns(neversink.df)

names(neversink.df)
```

# Plot the data
```{r}
neversink.plot <- ggplot(data=neversink.df,
                          aes(dateTime, Flow_Inst)) +
        geom_line()
neversink.plot
```


this is the interactive plot
```{r}
ggplotly(neversink.plot)
```



# Now read in *daily* data by year
```{r}
siteNo_daily <-     "01435000"
pCode_daily <-      "00060"
start.date_daily <- "1937-11-01"
end.date_daily <-   "2019-08-02"

nevversink_daily.df <-   readNWISdv(siteNumbers = siteNo_daily, 
                        parameterCd = pCode_daily,
                        startDate = start.date_daily, 
                        endDate = end.date_daily)
nevversink_daily.df <- renameNWISColumns(nevversink_daily.df)

names(nevversink_daily.df)

```


```{r}
daily.plot <- ggplot(nevversink_daily.df, aes(Date, Flow)) +
  geom_line()
daily.plot


```

```{r}
ggplotly(daily.plot)
```



```{r}
# siteNo_peak <-     "01435000"
# pCode_peak <-      "00060"
# start.date_peak <- "1850-01-01"
# end.date_peaky <-   "2019-08-02"
# 
# nevversink_peak.df <-   readNWISstat(siteNumbers = siteNo_peak,
#                         parameterCd = pCode_peak,
#                         startDate = start.date_peak,
#                         endDate = end.date_peak,
#                         statReportType="annual", 
#                         statType="mean")
  


```

